---
title: 'Reproducible Research: Peer Assessment 1'
author: "Paul Clark"
date: "December 25, 2016"
output: 
  html_document: 
    keep_md: yes
---
### Introduction

This assignment uses data from a personal activity monitoring
device. The data was collected from an anonymous individual during October and November, 2012, and includes the number of steps taken in 5 minute intervals. Here, we explore the data to:

 * Understand the distribution of steps per day
 * Profile the number of steps per 5-min period, averaged over the two months
 * Understand the impact of imputing missing values
 * Evaluate the differences between step profiles on weekdays vs. weekends

This document is structured so as to demonstrate fulfillment, in order, of the 9 requirements for a successful submission.  These are listed in the the section titled _**Commit containing full submission**_ within the assignment _**Review Criteria**_.    

***


#### 1. Set up environment and read data

Before starting, we must set up our environment and read the data.

```{r message = FALSE, warning = FALSE, collapse=TRUE, cache=FALSE}
# Set up our environment and directory structure
if (!require(dplyr)) {install.packages("dplyr")}
if (!require(ggplot2)) {install.packages("ggplot2")}
# Create directory for figures
if (!file.exists("figures")) {dir.create("figures")}
```
```{r cache=FALSE}
# Read in the data from the zipped file
conxn <- unz("activity.zip","activity.csv")
# Note: The file consists of 3 columns of data: steps, date, and 5-min interval.
activity <- read.csv(conxn, colClasses = c("integer", "character", "integer"))
# `read.csv` closes the conxn, but we also remove it from our environment.
rm(conxn)
```
***


#### 2. Histogram of number of steps taken per day

To obtain the histogram, we first group the `activity` dataframe by date and sum over the steps taken each day. Note that in doing so, it's very important to set `na.rm = FALSE` in `sum`, as otherwise the `summarize` function will include zeroes for dates with `NA`s. We want these dates to be properly recognized as containing missing data.

```{r cache= FALSE}
# Computing steps by date
stepsbydt_na <- activity %>% group_by(date) %>% summarize(StepsPerDay = sum(steps, na.rm = FALSE))
```

We then plot the desired histogram using `ggplot`, and save it in the target directory.  We let `na.rm = FALSE` so that explicit warning messages are generated, confirming the number of days for which the underlying data contained `NA`s.

```{r cache = FALSE, collapse = FALSE}
# Plotting the histogram
(histNoNA <- ggplot(stepsbydt_na, aes(StepsPerDay/1000)) + stat_bin(binwidth = 1, center = 0.5,
        closed = "right", alpha = 0.4, na.rm = FALSE) + 
        scale_x_continuous(minor_breaks = seq(0,22,1), labels = function(x){paste0(x, "K")}) +
        scale_y_continuous(breaks = seq(0,20,2)) +
        labs(x = "Thousands of Steps per Day", y = "Number of Days", title =
        "Histogram of Steps Taken per Day (Excluding NAs)"))
```
```{r warning=FALSE, message=FALSE, collapse=TRUE, results='hide'}
# Saving the histogram as a separate file
png("figures/histNoNA.png", width = 1000)
print(histNoNA)
dev.off()
```
***


#### 3. Mean and median steps per day

We compute the mean and median steps for data portrayed in the histogram using the `summary` function, which gives a bit more information than `mean` and `median`.  The median `r median(stepsbydt_na$StepsPerDay, na.rm = TRUE)` and mean `r format(round(mean(stepsbydt_na$StepsPerDay, na.rm = TRUE), 0), nsmall=0)` are quite close, consistent with a fairly symmetrical distribution.  Note that the above values were computed using `mean()` and `median()`, which give slightly different values than the `summary()` function.

```{r collapse=FALSE}
summary(stepsbydt_na$StepsPerDay, na.rm = TRUE)
```
***


#### 4. Time series plot of average number of steps

We use `dplyr` to group by interval, then average steps across all days for each interval. Here, we set `na.rm = TRUE` to eliminate `NA` results. We also normalize the "steps" value as average steps per minute by dividing by 5.
```{r}
stepsbyint <- activity %>% group_by(interval) %>% summarize(StepsPerMin = mean(steps, na.rm = TRUE)/5)
```
To make the plot as a function of time of day using `ggplot`, we use the interval information to generate times of class `POSIXct`. To do so, we need to reformat the interval information, which is read in as an integer that represents 24-hour clock time, not number of minutes. Note the break between 55 and 100, indicative of clock time (0:55 and 1:00):
```{r collapse=TRUE}
stepsbyint$interval[9:15]
```
The first step of our reformatting will convert these integers to character clock times.  To do that, we create a reformatting function and vectorize it so that we can pass it to `transmute`.
```{r}
# Definition of function to change interval into a character time of day
mktime <- function(tm) {
        digit4_tm <- paste0(paste0(rep("0", 4 - nchar(tm)), collapse = ""), tm, collapse = "")
        digit4_tm <- paste0(substr(digit4_tm, start = 1, stop = 2), ":", substr(digit4_tm, start = 3, stop = 4))
}
# Vectorize fn and use in transmute
vtime <- Vectorize(mktime, SIMPLIFY = TRUE, USE.NAMES = FALSE)
stepsbyintchar <- stepsbyint %>% transmute(time = vtime(interval), StepsPerMin)
```
Now the intervals are expressed as standard 24-hr clock times:
```{r collapse=TRUE}
head(stepsbyintchar)
```
Next, we convert to `POSIXct` and create the plots (one for display, one for safe-keeping). Note that all times will of necessity be accorded a date, as required by `POSIXct`. However, the date is of no consequence, so long as we do not use a day with changes to/from Daylight Savings Time!  We prepare the variable `plottime` by appending the first date of the period in question, on which we know no conversion to/from DST occurs:
```{r cache=FALSE, results ='hide'}
stepsbyintchar$plottime <- paste("2012-10-01", stepsbyintchar$time)
# convert character time to actual dttm (POSIXct) value
stepsbyinttm <- stepsbyintchar %>% transmute(plottime = as.POSIXct(plottime, format = "%Y-%m-%d %H:%M"), StepsPerMin)
# plot average number of steps taken over course of day, with time shown in clock hours
(stepTimeSeriesAvg <- ggplot(stepsbyinttm, aes(x = plottime, y = 5*StepsPerMin)) + geom_line() +
        scale_x_datetime(date_labels = "%I %p") + labs(x = "Time of Day",
        y = "Average Steps Per 5-Min Interval", 
        title = "Average steps per 5-min interval throughout day"))
# Save time series plot in figures directory
png("figures/stepTimeSeriesAvg.png", width = 1000)
print(stepTimeSeriesAvg)
dev.off()
```
***


#### 5. The 5-minute interval that contains the maximum average number of steps
```{r}
(time_maxsteps <- stepsbyintchar[which.max(stepsbyintchar$StepsPerMin),])
```
The time is at `r time_maxsteps$time[[1]]` on the 24-hour clock.  

***


#### 6. Impute missing data

The total number of missing values in the data set is given by the last table value of the `summary()` function, when `na.rm = FALSE` (of course, we could also use `is.na()`, but `summary()` gives more information):
```{r collapse = FALSE}
summary(activity$steps, na.rm = FALSE)
```
Note that this is exactly 8 days of missing data. To show that, we divide the total number of missing values by 24 hours/day * 12 intervals per hour, or `r 12*24` intervals per day:
```{r collapse = TRUE}
(eight <- sum(is.na(activity$steps))/(12*24))
```
The `r eight` days in question are the following:
```{r collapse = TRUE}
with(activity, unique(date[is.na(steps)]))
```

We now create a new dataset with these missing values filled in. We proceed by replacing each missing value with the value for the associated 5-minute interval computed by averaging across all days.  Since we know the exact location of all the `NA`s, we could replace the values rather simply.  Instead, we use a more general procedure by following these steps:  

  1.  Use `is.na()` to create a vector of `r sum(is.na(activity$steps))` interval values associated with each of the NAs.  
  2.  Use `match()` to match those interval values to those in the dataframe of average steps vs. interval. This generates a vector of indices (row numbers) for those intervals in the average steps dataframe.  
  3.  Use the indices to 'look up' the correct average value for the interval associated with each `NA` and, using `is.na()` again, replace it in the original data set.
```{r collapse=TRUE}
# Create vector of interval values needing imputation
NAintervals <- activity[is.na(activity$steps), "interval"]
# Construct vector containing the corresponding indices in the `stepsbyint` dataframe
NAindices <- match(NAintervals, stepsbyint$interval) # Vector of length = number of NAs
# Use the index values in 'NAindices' to replace the NA values in the orginal data-set
# with those in the mean step per interval dataframe. (Multiply by 5, since we normalized
# the interval data in 'stepsbyint' to a value in average steps per minute).
activityimp <- activity
activityimp$steps[is.na(activity$steps)] <- 5*(stepsbyint$StepsPerMin[NAindices])
```
***


#### 7. Make a histogram of steps-per-day after imputing NAs, reporting on changes in mean and median

As in section (2), we group our data by date and sum over steps. Since there are now no `NA`s, we let `na.rm = FALSE` in `sum()`, so that a warning is generated in the event we have missed any.
```{r}
stepsbydtimp <- activityimp %>% group_by(date) %>% summarize(StepsPerDay = sum(steps, na.rm = FALSE))
```

The new mean and median number of steps taken each day:
```{r}
summary(stepsbydtimp$StepsPerDay, na.rm = FALSE)
```
The difference between `mean` and `median` is now `r mean(stepsbydtimp$StepsPerDay) - median(stepsbydtimp$StepsPerDay)`, consistent with a distribution even more symmetrical than before.  We now calculate the difference between imputed and non-imputed data on mean and median.  We must remove the `NA`s to obtain the mean and median for the data with `NA`s.
```{r collapse=TRUE}
mean(stepsbydtimp$StepsPerDay) - mean(stepsbydt_na$StepsPerDay, na.rm = TRUE)
median(stepsbydtimp$StepsPerDay) - median(stepsbydt_na$StepsPerDay, na.rm = TRUE)
```
By adding the imputed values, the mean did not change at all.  This is to be expected, since
all `r eight` days with `NA` values had imputed means equal to the population mean prior to imputing NAs.  However, the median Steps Per Day moves up slightly, since the original mean, `r format(round(mean(stepsbydt_na$StepsPerDay, na.rm = TRUE), 0), nsmall=0)`, is slightly higher than the original median, `r median(stepsbydt_na$StepsPerDay, na.rm = TRUE)`. (Note: there is a disparity between the mean as reported via the `mean()` function, in this paragraph, and that reported by the `summary()` function, above.  We have run various experiments to see what might be causing this difference, but have not yet reached a conclusion.) 

We can see the precise change to the distribution most clearly if we do a 2-panel plot, juxtaposing histograms before and after imputing `NA` values.  To do this, we first compute a factor variable that will be used to differentiate the two histograms in `ggplot`, then combine the two dataframes (with and without imputed data), then call `ggplot`.  This time, `ggplot()` warns us again, as last time, that there are `r eight` missing values (days) in the pre-imputed data set.
```{r results='hide'}
# Create factor variable
NA_Treatment <- as.factor(c(rep("NAs Excluded", nrow(stepsbydt_na)), rep("NAs Imputed", nrow(stepsbydtimp))))
# Combine the two dataframes
stepsbydt_impexc <- rbind(stepsbydt_na, stepsbydtimp)
# Add factor variable to dataframe
stepsbydt_impexc$NA_Treatment <- NA_Treatment
# Make 2-panel plot
(histPanelNAimputed <- ggplot(stepsbydt_impexc, aes(StepsPerDay/1000)) + stat_bin(binwidth = 1,
        center = 0.5, closed = "right", alpha = 0.4, na.rm = FALSE) + 
        scale_x_continuous(minor_breaks = seq(0,22,1), labels = function(x){paste0(x, "K")}) +
        scale_y_continuous(breaks = seq(0,20,4)) +
        facet_grid(NA_Treatment~.) +
        labs(x = "Thousand Steps Per Day", y = "Number of Days", title =
        "Histograms by Treatment of NAs: Number of Days per Thousand Steps"))
```
```{r results = 'hide', collapse=TRUE}
# Save figure to disk, for safe-keeping
png("figures/histPanelNAimputed.png")
print(histPanelNAimputed)
dev.off()
```
***


#### 8.  Panel plot comparing average number of steps per interval across weekdays and weekends

Creation of the panel plot takes multiple steps, but is a lot like the process in (4) above. However, we need to take extra steps to compute separate averages by weeekday and weekend day. Basically, we convert the character `date` field of our data set to `POSIXct` format, so that we can call `weekdays()` on it. Then we are in a position to compute the required averages, this time for both weekdays and weekend days.
```{r}
# As in (4), convert 'date' column of 'activityimp' dataframe to POSIXct format
activityimp$posixdate <- with(activityimp, as.POSIXct(date))
# From 'posixdate' column, create a 'dayofwk' column
activityimp$dayofwk <- with(activityimp, weekdays(posixdate, abbreviate = TRUE))
# From the 'date' column, use fn ifelse() to create colum 'daytype' containing 
# factor variable with two levels: 'weekday' and 'weekend'
activityimp$daytype <- with(activityimp, as.factor(ifelse(dayofwk == "Sat" | dayofwk == "Sun", "weekend","weekday")))
# Use dplyr to simultaneously call 'group_by' on the 'activityimp' dataframe with arguments
# 'interval' and 'daytype', then call 'summarize' on 'steps' using the `mean()` function.
stepsbyintdytype <- activityimp %>% group_by(interval, daytype) %>% summarize(StepsPerMin = mean(steps, na.rm = TRUE)/5)
```
The remaining steps are just like those in (4), with the exception of the call to `facet_grid()` with `ggplot()`.
```{r results='hide'}
# Leverage previous function 'mktime' to change intervals into character times of day.
# In particular, leverage vectorized form of fn, 'vtime',  for use in dplyr 'transmute'.
stepsbyintchardytype <- stepsbyintdytype %>% as.data.frame %>%
                transmute(interval, time = vtime(interval), StepsPerMin, daytype)
# We prepare the variable 'plottime' by appending the first date of the period in question
stepsbyintchardytype$plottime <- paste("2012-10-01", stepsbyintchardytype$time)
# Note: 'as.date.frame()' needed to remove grouping information prior to transmuting
stepsbytmdytype <- stepsbyintchardytype %>% as.data.frame %>% transmute(plottime =
                as.POSIXct(plottime, format = "%Y-%m-%d %H:%M"), StepsPerMin, daytype)
# Plot average number of steps taken per 5-Min period in panel plot comparing 
# weekday and weekend steps
(stepTimeSeriesPanel <- ggplot(stepsbytmdytype, aes(x = plottime, y = 5*StepsPerMin)) + 
        facet_grid(daytype ~.) + geom_line() +
        scale_x_datetime(date_labels = "%I %p") + labs(x = "Time of Day",
        y = "Average Steps Per 5-Min Interval",
        title = "Average steps per 5-min interval throughout day: Weekdays vs. Weekends"))
# Save a version of the plot in the figures directory
png("figures/stepTimeSeriesPanel.png", width = 800)
print(stepTimeSeriesPanel)
dev.off()
```
By eye-balling the charts, one can see that on weekends (as compared to weekdays) our walker
is slower reaching maximum speed, has a lower maximum speed, and is active later in the day (the distribution of steps has its final local maximum around 8 PM or so on weekends vs. 7 PM on weekdays).  

***


#### 9. All code used in analysis

All code used in this analyis is visible in this document.  

***

### Appendix: Session Info for Reproducibility of Results

```{r}
sessionInfo()
```
